Bootstrapping Multi-atlas Hippocampal Segmentation with MAGeT

### 2013-02-11 paper.Rnw
Some of the methods rewritten for clarity

Fully-automated segmentation techniques require no human expertise beyond that
of initial setup or training. One broad class of automated approach is that of
the {\em model-based} approaches which employ models of anatomical feature
variation to constrain segmentations. In this paper we focus on a subclass known
as {\em multi-atlas} approaches which make use of a series of expertly segmented
MRIs (the atlas library) as models to guide the segmentation of a query
subject's MRI, and achieve some of the best automated hippocampal segmentation
accuracies to-date.

The multi-atlas segmentation technique was pioneered by Heckemann et al. with
the MAPER algorithm \citep{Heckemann2006,Heckemann2011}. The basic strategy is as
follows. Each atlas image is fit a subject's neuroanatomy using nonlinear
registration techniques \citep{Collins1995,Klein2009}. This resulting nonlinear
transformation is then applied to the atlas labels to bring them into the
subject space.  Finally, a {\it label fusion} technique, such as voxel-wise
voting, is used to merge these labels into a definitive segmentation for the
subject. Typically, an atlas selection approach is used to improve accuracy by
only using labels from atlases that are most similar to a subject image
(cross-correlation of image intensities is a common similarity metric)
\citep{Aljabar2009}.

% - Warfield2004: STAPLE as an alternative decision fusion technique?

Multi-atlas techniques have been applied very successfully to hippocampal
segmentation. Collins et al. found that with an atlas library of 80 atlas
images, the use of the ANIMAL nonlinear registration algorithm, the normalised
mutual information as similarity metric for atlas selection, and majority vote
for decision fusion they were able to achieve a mean Dice Kappa similarity score
(DSC) to manual labels of 0.886\citep{Collins2010}. The Alzheimer's Disease
Neuroimaging Initiative (ADNI) is a commonly used benchmarking dataset of MR
images of controls and patients with MCI or Alzheimer's (see Methods for more
information on the ADNI dataset).  Leung et al. tuned a multi-atlas  approach to
the segmentation of ADNI1 1 year dataset of images and using 55 atlases and the
STAPLE decision fusion method \citep{Warfield2004} achieved an mean DSC of 0.89
to the manual segmentations supplied with the ADNI data  \citep{Leung2010}. The
MAPER whole brain segmentation algorithm (described above), using 30 atlases,
achieved a mean DSC score of 0.889 on all ADNI1 baseline images
\citep{Heckemann2011}. In \citep{Lotjonen2012}, the authors use a proprietary
non-linear registration method based on intensity differences, and
post-processing step taking into account these intensity differences to tidy up
fused segmentations. Using images from the ADNI1 baseline dataset, they achieved
a mean DSC of 0.885 with 30 atlases. With the LEAP algorithm, Wolz et al.
explored an elegant modification to the basic strategy: the atlas library is
grown, beginning with a set of manually labelled atlases and successively
incorporating unlabelled subject images after being labelling using multi-atlas
techniques\citep{Wolz2010}.  The sequence in which subject images are labelled
is chosen so that the similarity between the atlas images and the target images
is minimised at each step, effectively allowing for deformations between very
dissimilar images to be broken up into sequences of smaller deformations.  With
an atlas library of 30 MR images, LEAP was used to segment the ADNI1 baseline
dataset, achieving a mean Dice score of 0.85 with manual segmentations.

@@

\subsection{MAGeT Brain Algorithm}
\todo{Make sure we explain what MAGeT stands for}
\todo{reference Chakravarty2012}

In this paper, we use the term {\it atlas} to mean any manually segmented MR
image, and the term {\it atlas library} to mean a set of such images.  We use
the term {\it template} to refer to any MR image, and associated labelling,
used to segment another image, and the term {\it template library} to refer to
a set of such images.  An atlas library may be used as a template library but,
as we will discuss, a template library may also be composed of images with
computer generated labellings.

The simplest form of multi-atlas segmentation combines labellings derived from
several atlases by way of label fusion \citep{Heckemann2006}.  We will refer to
this as {\em basic multi-atlas segmentation}.  The primary steps are as
follow are: (1) an atlas library and unlabelled MR images are given as input,
(2) each atlas image is nonlinearly registered to each unlabelled image, (4)
each atlas' labels are propagated via the resulting transformations to the
unlabelled image space, and (5) the resulting labels are fused to produce a
single, definitive segmentation. The particular registration and voting method
used are left unspecified.

The segmentation approach we propose is best understood as an extension of
multi-atlas segmentation.  MAGeT brain adds a preliminary stage to this process
in which the template library is constructed rather than given as input.  As
before, MAGeT brain accepts an atlas library and unlabelled MR subject images as
input.  Template library images are selected from subject images.  The choice of
subjects used in the template library can be made to reflect the neuroanatomy or
demographics of the subject set as a whole (for instance, by sampling equally
from cases and controls).

To create the template library, labels from each atlas image are propagated to
each template library image via the transformation resulting from a non-linear
registration between pair of images. As a result, each template library image
has a label from each atlas. Basic multi-atlas segmentation is then used to
produce segmentations for the entire set of unlabelled images (including those
images used in the template library). Figure \ref{alg:MAGeT} describes the MAGeT
brain algorithm in pseudocode.  Source code can be found at
\url{http://github.com/pipitone/MAGeTbrain}.

>>>>

Automated segmentation techniques do not require human intervention but do
require {\it a priori} anatomical information to guide segmentations.  In this
paper we focus on methods that use manually segmented atlases as anatomical
priors, as these methods achieve some of the best automated hippocampal
segmentation accuracies to-date. This technique was pioneered by []\citep{} with
single-atlas (or model-based) segmentation.  Volumetric image registration is
used to estimate a fit between the neuroanatomy of an atlas and subject images.
Labelling of the subject image is achieved by applying the resulting
transformation to the atlas labels to bring them into the subject image space
(\{em label propagation}). This method is limited in accuracy by the
introduction of {methodological} errors in registration and partial volume
effects in label resampling, and {conceptual} errors introduced when the anatomy
of the atlas is not representative of the subject.

Multi-atlas segmentation techniques address these limitations by combining
segmentation information from a series of expertly segmentated atlases
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2012,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a subject image, and label propagation is
performed to produce several labellings of the subject image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used merge
these labels into a definitive segmentation for the subject.

{\em Atlas selection} techniques are often used to filter out atlases dissimilar
to a subject image in order to reduce {conceptual} errors \citep{Aljabar2009}.

Multi-atlas methods have been very successfully applied to hippocampal
segmentation. Collins et al. found near-manual segmentation performance using an
atlas library of 80 T1-weighted atlas images from the ICBM152 dataset, the
ANIMAL nonlinear registration algorithm, normalised mutual information as a
similarity metric for atlas selection, and majority vote for label fusion
\citep{Collins2010}.

The Alzheimer's Disease Neuroimaging Initiative (ADNI) is a
commonly used benchmarking dataset of MR images of controls and patients with
MCI or Alzheimer's (see Methods for more information on the ADNI dataset).

Leung et al. tuned parameters for registration and label fusion to the
segmentation of ADNI1 1 year dataset of images with an atlas library of 55
images \citep{Leung2010}.

The MAPER whole brain segmentation algorithm \citep{Heckemann2006,Heckemann2011},
using 30 atlases, on all ADNI1 baseline images \citep{Heckemann2011}.

In \citep{Lotjonen2012}, the authors use a proprietary non-linear registration
method based on intensity differences, and post-processing step taking into
account these intensity differences to 
- TODO: explain "tidy up"
tidy up
fused segmentations. Using images from the ADNI1 baseline dataset, they achieved
a mean DSC of 0.885 with 30 atlases.

The LEAP algorithm, Wolz et al.  is an elegant modification to the basic
multi-atlas strategy: the atlas library is grown, beginning with a set of
manually labelled atlases and successively incorporating unlabelled subject
images after being labelling using multi-atlas techniques\citep{Wolz2010}.  The
sequence in which subject images are labelled is chosen so that the similarity
between the atlas images and the target images is minimised at each step,
effectively allowing for deformations between very dissimilar images to be
broken up into sequences of smaller deformations.  With an atlas library of 30
MR images, LEAP was used to segment the ADNI1 baseline dataset, achieving a mean
Dice score of 0.85 with manual segmentations.

- TODO: "probablistic methods?"

@@

\subsection{MAGeT Brain Algorithm}
We will use the term {\it atlas} to refer to two image volumes: an intensity
image ({\it atlas image}) and a corresponding manual segmentation image ({\it
atlas labels}). A {\it subject} image is an intensity image for which we would
like an segmentation.  The term {\it template} will be used to refer more
generically to any image and corresponding labelling, manual or computed, when
it is {\it used as} as a model in the segmentation of another image. In other
publications, the terms atlas and template are used interchangeably, but in this
paper we force the distinction to make it clear when an image has a manual
segmentation.  The terms {\it atlas library} and {\it template library} mean a
set of such images.

The simplest form of multi-atlas segmentation combines labellings derived from
several atlases by way of label fusion \citep{Heckemann2006,Heckemann2011}. We
will refer to this as {\em basic multi-atlas segmentation}, and specific the
basic schema as:

    1. an atlas library and set of subject images are given as
       input, and the atlas library is used as a template library in the
       following steps; 
    2. each atlas intensity image is nonlinearly registered to each subject
       intensity image; 
    3. label images from each atlas are propagated via the resulting
       transformations to the subject image space; and 
    4. the resulting labels are fused to produce a single, definitive
       segmentation.

The particular registration and voting method used are left unspecified.

MAGeT brain is best understood as an extension of
the basic multi-atlas segmentation schema. An additional stage is included in
which the template library is constructed by labelling a subset of subject
images, rather than by using the atlas library. The choice of subjects used in
the template library can be made to reflect the neuroanatomy or demographics of
the subject set as a whole (for instance, by sampling equally from cases and
controls).

Labels from each atlas image are propagated to
each template library image via the transformation resulting from a non-linear
registration between pair of images. As a result, each template library image
has a label from each atlas. Basic multi-atlas segmentation is then used to
produce segmentations for the entire set of unlabelled images (including those
images used in the template library). Figure \ref{alg:MAGeT} describes the MAGeT
brain algorithm in pseudocode.  Source code can be found at
\url{http://github.com/pipitone/MAGeTbrain}.
